On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-17 manuscripts/PNAS/chazthm_pnas.tex
discussion

% A significant implication is that the constant $C_1$ of Thm.~\ref{DeterministicUniquenessTheorem} can be computed efficiently if we restrict the class of possible supports for the $\mathbf{\bar x}_i$.

In this note, we prove the stability of unique solutions to Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} while significantly reducing the known sample complexity. Our results justify the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, as in \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research. The main motivation for this work, however, was to understand how seemingly universal representations emerge from various sparse coding models fit to natural data.

What we have shown here is that the sparse linear coding model generally produces a \textit{well-posed} inverse problem to be approximately solved by a numerical algorithm. This early concept of Hadamard \cite{Hadamard1902} can be paraphrased as the idea that inferences from observations should be robust to the inevitable uncertainty in measurement. In other words, a small perturbation of the data should result in only slightly different inferred parameters. In this regard, for the sparse linear coding model we demonstrate a linear relationship, \eqref{epsdel}, between measurement noise and the deviation of any solution from the true parameters, with explicit constants expressed in terms of these parameters. Moreover, we show that even if the meta-parameter for the number of dictionary elements is overestimated, a subset of parameters may still be identifiable up to noise. We remark that our constants have been derived for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be smaller when sampled from a distribution.

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singleton intersection property, Def.~\ref{sip}) for the support sets of sparse codes key to the identification of the dictionary. Fully understanding those combinatorial designs allowing for stable sparse representations is an interesting research area for the future. For instance, whether there is a sample size $N$ that is polynomial in $m,k$ gauranteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} and related questions \cite{Tillmann15}. A reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} demonstrates that this can in fact be achieved if one restricts the supports of the $\mathbf{\bar x}_i$ to a class that grows polynomially in $m, k$.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems.

In the field of theoretical neuroscience in particular, dictionary learning for sparse coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that some of the reproduce response properties of cortical neurons. Our results suggest that this correspondence could be due to the ``universality'' of sparse representations in natural data, an early idea in neural theory \cite{pitts1947}. %Furthermore, they justify the hypothesis of \cite{Isely10} that sparse codes passed through information bottlenecks in the brain are recovered from random projections via (unsupervised) biologically plausible sparse coding (e.g., \cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}).
Deep learning noise issues: \cite{goodfellow2014explaining} and neuroscience theory \cite{ganguli2012compressed}.

>>>>

% A significant implication is that the constant $C_1$ of Thm.~\ref{DeterministicUniquenessTheorem} can be computed efficiently if we restrict the class of possible supports for the $\mathbf{\bar x}_i$. 
In the last several decades, from JPEG's dominance in image compression to the discovery of intrinsic ``Gabor'' wavelets in natural images \cite{donoho2001can} to the detection of forgeries \cite{hughes2010, Olshausen10} to artificial ``cat detecting neurons'' in machine intelligence, sparse linear coding has become a standard tool in signal analysis, allowing for the unsupervised discovery of myriad insights into natural signals. 
%A compelling feature of this model is its simple instantiation of the principle of Occam's razor: a natural video, for instance, is modeled as a linear combination of a small number of spatiotemporal building blocks, each representing an archetypical feature latent in the data. 
In the field of theoretical neuroscience in particular, dictionary learning for sparse linear coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that reproduce some response properties of certain cortical neurons. The motivation for our work here was to understand when such universal representations can be expected to emerge from the sparse linear coding model.

%We have demonstrated the uniqueness and stability of solutions to Prob. \ref{InverseProblem} and the related Prob. \ref{OptimizationProblem}, 
To this end, we have conclusively verified that inference in this parsimonious model of (noisy) natural data is a well-posed problem in the sense of Hadamard \cite{Hadamard1902}, a long-standing necessary property of reductionist models in science. This notion summarizes the idea that inferences from observations should be robust to the inevitable uncertainty in measurement. In this regard, we demonstrate rigorously that, given sufficient data and suitable dictionary, Prob.~\ref{OptimizationProblem} reduces to an instance of Prob.~\ref{InverseProblem} to which applies our main result (Thm.~\ref{DeterministicUniquenessTheorem}): every dictionary and sparse codes solving Prob.~\ref{InverseProblem} are equivalent up to inherent relabeling/scaling ambiguities and a discrepancy (error) that scales linearly with the measurement noise or modeling inaccuracy; and we provide explicit constants expressed in terms of these latent parameters.
%In other words, a small perturbation of the data should result in only slightly different inferred parameters. 
%Moreover, we show that even if the meta-parameter for the number of dictionary elements is overestimated, a subset of parameters may still be identifiable up to noise. 
%This work justifies the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, as in \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research.

An immediate application of our theoretical work is Prop.~\ref{neuroprop}, which certifies the validity of perhaps the only published theory of neurally-plausible bottleneck communication between brain regions -- that unsupervised feature (dictionary) learning in a significantly smaller space of nervous activity can recover noisy sparse codes sent through a randomly-constructed (but unknown) wiring bottleneck \cite{Isely10}.  We refer the reader to \cite{ganguli2012compressed} for a discussion of this application's significance and more on the interactions between sparse linear coding and neuroscience.

One of the major innovations in our work is a theory of combinatorial designs for support sets (the ``singleton intersection property'') that provides a handle on the complexities of intersecting subspaces in dictionary learning problems. We incorporate this idea into a new fundamental lemma in combinatorial matrix theory (Lem.~\ref{MainLemma}) that will be useful in any noisy setting where uniqueness up to feature relabelling/scaling is desirable, and which draws upon the definition of a new matrix lower bound induced by a given hypergraph. Our novel combined combinatorial and analytical proof allows for the following new theoretical advances over previous works (e.g. \cite{Hillar15}) beyond an extension to the noisy case: 1) the number of dictionary elements can be unknown; 2) data require a polynomial, not exponential, number of distinct sparse supports and in certain settings a practical (polynomial) number of samples in total is sufficient; and 3) the spark condition is not necessary for recovery of the dictionary.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems. To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these multi-layer feedforward models of data. Moreover, it is now well-known in the community that even minuscule amounts of (``adversarial'') signal noise can dramatically alter their output \cite{goodfellow2014explaining}. In particular, it appears that a main reason for the sustained interest in unsupervised (sparse) dictionary learning for data analysis has been the assumed well-posedness of the model, confirmation of which forms the core of our  findings.

 [*** TODO: PUT in paper sprinkled TIGHTNESS and COMPUTABILITY ***]

%We make a final remark about the tightness of our results and the computability of our derived constants.
%We remark that our constants have been derived for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be smaller when sampled from a distribution.

%Our results suggest that this correspondence could be due to the ``universality'' of sparse representations in natural data, an early idea in neural theory \cite{pitts1947}.

%A reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} demonstrates that this result enables a practical amount of data to determine the model if one restricts the supports of the $\mathbf{\bar x}_i$ to a class that grows polynomially in $m, k$.
